Train Harder, Smarter:\\Using Graph Theory to Create Route Suggestions

### 2013-11-08 paper.tex
Added the pyroute files to the repo.
This is perhaps a bit silly, as most of the development will
occur only in the pyroutelib functions, but it's being done
nonetheless.

\setlength{\parskip}{5pt plus1pt minus1pt}
\setlength{\headheight}{15pt}

@@

\title{Train Harder, Smarter:\\Using Graph Theory to Create Route Sugestions}
\author{Forest Trimble\\trimbf@rpi.edu}
\begin{document}

@@

\begin{abstract}
  \emph{Cyclists are always in search of the perfect ride on the perfect
  roads. They have criteria like distance and elevation gain to ensure
  that they get in the workout that they want. Unfortunately, in order to
  find this ideal ride manually, it takes a great deal of exploring and
  time, and eventually, one may settle into the habit of using the same
  roads that he/she already knows. We research a way to improve this
  paradigm, and to generate cyclists exactly the route they are looking
  for, without them having to do any work.}
\end{abstract}

@@

Specifically, a good solution should take as input a distance to travel, a
start point, and, optionally, an elevation gain, and generate a route from the
start point that satisfies the distance and elevation gain within some
$\epsilon$. A fully robust solution would also utilize user data to ensure that
it traverses the most pleasant roads to ride on and avoids the worst.

@@

Creating a solution to this problem from scratch would require immense amounts
of time and effort that are simply not available. Fortunately, the larger part
of this problem has already been solved: maps have already been digitized and
parsers for the data format already exist. This section details the various
tools and technologies that the algorithm leverages.

@@

INFO ON OSRM

@@

The most important part of all of this is how it actually works. The core idea
is that at any point along the route, the engine generates weights for each
different possible direction to take, generates a random number, and picks a
direction according to the random number and the weights. There are several
different ideas that are competing for weight, so we'll break it down into the
individual components first, and then describe the combination at the end.

@@

In order to direct you along this loop, weights are generated based on
direction. We'll consider the method for finding the optimal direction,
$\theta$, briefly. First, consider how the weights are generated once
that direction is discovered. Basically, there is a weighted compass oriented
along $\theta$, with weights from 0.5 to 2. See Figure~\ref{fig:weights} to
see what we mean. As shown, the weights do not scale linearly, instead, the
weight of any angle, $\tau$, is calculated as follows:
\begin{equation}
  \mbox{WEIGHT} = 2^{1-\frac{2}{\pi}|\theta-\tau|}. \label{eq:weight}
\end{equation}

@@

ARE THERE MORE FOR THIS SECTION?

@@

>>>>

\setlength{\parskip}{5pt plus1pt minus1pt}

@@

\title{Train Harder, Smarter:\\Using Graph Theory to Create Route Suggestions}
\author{Forest Trimble\\trimbf@rpi.edu}
\begin{document}

\setlength{\headheight}{15pt}

@@

\begin{abstract}
  \emph{Cyclists are always in search of the perfect ride on the perfect
  roads. They have criteria like distance and elevation gain to ensure
  that they get in the workout that they want. Unfortunately, to
  find this ideal ride manually, it takes a great deal of exploring and
  time, and eventually, one may settle into the habit of using the same
  roads that he/she already knows. We research a way to improve this
  paradigm, and to generate cyclists exactly the route they are looking
  for, without them having to do any work.}
\end{abstract}

@@

Specifically, a good solution should take as input a distance to travel, a
start point, and, optionally, an elevation gain, and generate a route from the
start point that satisfies the distance and elevation gain within some
tolerance. A fully robust solution would also utilize user data to ensure that
it traverses the most pleasant roads to ride on and avoids the worst.

@@

Creating a solution to this problem from scratch is unnecessary. The larger part
of this problem has already been solved: maps have been digitized and
parsers for the data format already exist. This section details the various
tools and technologies that the algorithm leverages.

@@

Additionally, we leveraged some existing code to more easily be able to create
a demonstrable algorithm. Pyroute was used, since it has the very nice feature
that the navigation component was entirely decoupled into PyrouteLib. This
allows easy utilization of the Pyroute tools for GUI setup and OSM download
while also making it rather explicit where the navigation functions are. The
OSM wiki pages have a good deal of information
on Pyroute and PyrouteLib; refer to them for more information.

@@

The core idea of our algorithm
is that at any point along the route, the engine generates weights for each
different possible direction to take, generates a random number, and picks a
direction according to the random number and the weights. There are several
different ideas that are competing for weight, so we'll break it down into the
individual components first, and then describe the combination at the end.

@@

In order to direct you along this loop, weights are generated based on
direction. We'll consider the method for finding the optimal direction,
$\theta$, briefly. First, consider how the weights are generated once
that direction is discovered. Basically, there is a weighted compass oriented
along $\theta$, with weights from 0.5 to 2. See Figure~\ref{fig:weights} to
see what we mean. As shown, the weights do not scale linearly. Instead, they
scale exponentially. This ensures that the weights are mostly centered around
the perpendicular, with growth accelerating farther from the perpendicular.
The idea is to have the weight be double in the ideal
direction, half moving exactly opposite the ideal direction, and unchanged
perpendicular to the ideal direction. Considering these three pieces of
information, we are looking for some $f : [0,2\pi] \to [0.5,2]$ as follows:
\begin{align*}
  2 = & f(\theta) \\
  1 = & f(\theta \pm \frac{\pi}{2}) \\
  \frac{1}{2} = & f(\theta \pm \pi)
\end{align*}
Let $\tau$ be the angle under consideration. We let
\[ \mbox{ang}(\tau, \theta) = |\theta - \tau \mbox{ mod } \pi|. \]
This allows us to define $f$ in terms of the angle between $\tau$ and
$\theta$. This function is necessary because $\theta$ is not necessarily
0. Note that this function only allows $\mbox{ang}(\tau, \theta)
\in [0,\pi]$. We accept this since we would like a compass symmetric
over the angle $\theta$, and we don't care if the angles are negative
or positive. However, the original function was defined in terms of the
value of $\tau$; instead, we are interested in a function based on the
angle between $\tau$ and $\theta$, which requires a slight redefinition:
\begin{align*}
  2 = & g(0) \\
  1 = & g(\frac{\pi}{2}) \\
  \frac{1}{2} = & g(\pi)
\end{align*}
We've already expressed the fact that we're interested in an exponential
function, and a base of two seems the obvious choice, leading to another
reduction:
\begin{align*}
  2 = 2^{h(0)} & \to & h(0) = 1 \\
  1 = 2^{h(\frac{\pi}{2})} & \to & h(\frac{\pi}{2}) = 0\\
  \frac{1}{2} = 2^{h(\pi)} & \to & h(\pi) = -1
\end{align*}
This yields two sets of three collinear points, which is easy enough to
map one to the other. Consider $h : [0,\pi] \to [-1,1]$ as follows:
\[ h(\mbox{ang}(\tau,\theta)) = 1 - \frac{2}{\pi}\mbox{ang}(\tau, \theta). \]
This satisfies the three points we gave, so we plug it in and calculate weight
accordingly:
\begin{equation}
  \mbox{WEIGHT} = 2^{1-\frac{2}{\pi}\mbox{ang}(\tau,\theta)}. \label{eq:weight}
\end{equation}

@@

@@

\section{Experimentation \& Results}

\section{Discussion}

\section{Future Work}

There are many ways to expand the algorithm developed here. While we have done
our best to utilize as much data as possible, expanding the set of data utilized
to generate the routes will likely improve the algorithm.

 (ASK DAN FOR THIS
PAPER) There has been research done into analyzing the probability of accidents
on a road based on certain metadata. This metadata is readily available to the
cities that own the roads, but it is seldom available for direct access inside
of maps. If this data becomes more readily available, then it can be integrated
into the application to help increase user safety.

Additionally, this has been a facet of a three pronged project developing a
platform for bicycle training. The other two aspects are developing an Android
application to track training data during a ride and a website to track
long-term training data. Ideally, this algorithm will eventually be seamlessly
integrated into the mobile and web projects, so that routes can be generated
and sent directly to the application for the user to follow.

Once the three prongs of this project have been integrated together, leveraging
the data from the other parts would also provide valuable input for the
algorithm. One of the most exciting applications of this information is to
generate something similar to a heatmap of roads that users of the website
ride. With this heatmap, the weights could be improved and generated based on
popularity.

Finally, it might be interesting to generate routes that are not necessarily
circuits. Loops are generally more useful since cyclists are likely to want
to return home after a good training ride, but it might also be useful to
have some sort of quasi-navigation that aims for the most fun route
between two locations rather than the most direct route.
